Linear Ordered Attribute Grammar scheduling using SAT-solving

### 2014-10-17 paper/paper.tex
Optimisations

\section{Optimisations} \label{sect:optimisations}
We have shown that expressing the scheduling problem as a SAT problem and using an existing SAT-solver can improve the running time of the scheduling, but that is not the only advantage. In the SAT problem one can easily add extra constraints to further influence the resulting schedule. In this section we show two of such optimisations that are useful from an attribute grammar perspective. These optimisations have not been implemented in the release version of the UUAGC, but we have run preliminary experiments to verify that they work as expected.

\subsection{Minimising visits}
The result of the static scheduling is a runtime evaluator that computes the values of the attributes for a given abstract syntax tree. The total order for each nonterminal defines in what order attributes should be computed, but in the implementation of the evaluator we make use of a slightly bigger unit of computation: a \emph{visit}.

@@

It is theoretically impossible to minimise the total number of visits performed for the full abstract syntax tree, because at compile-time we do not have a concrete abstract syntax tree at hand and only know about the grammar. We therefore try to minimise the maximum number of visits that any nonterminal takes.

\Jeroen{todo: describe how}

\subsection{Eager attributes}
Another optimisation is the ability to define \emph{eager attributes}. Eager attributes are attributes that should be computed as soon as possible, and must be annotated by the attribute grammar programmer as such. We would like our scheduling algorithm then to schedule them as early as possible in the total order.

@@

\Jeroen{todo: describe how}

>>>>

\section{Optimisations} \label{sect:optimisations}
We have shown that expressing the scheduling problem as a SAT problem and using an existing SAT-solver can improve the running time of the scheduling, but that is not the only advantage. In the SAT problem one can easily add extra constraints to further influence the resulting schedule. In \sectref{sect:minimising} and \sectref{sect:eagerattributes} we show two of such optimisations that are useful from an attribute grammar perspective. These optimisations have not been implemented in the release version of the UUAGC, but we have run preliminary experiments to verify that they work as expected.

\subsection{Interacting with the solver}
Instead of directly expressing all constraints in the initial SAT problem, we use a different trick for implementing the two optimisations: interacting with the solver. After the initial scheduling problem has been solved, we can ask for the truth value of all variables to construct the schedule. MiniSat also keeps some state in memory that allows us to add extra constraints to the problem, and ask for a new solution. In this way we can start with an initial solution and interact with the solver until some optimum has been reached.

\subsection{Minimising visits} \label{sect:minimising}
The result of the static scheduling is a runtime evaluator that computes the values of the attributes for a given abstract syntax tree. The total order for each nonterminal defines in what order attributes should be computed, but in the implementation of the evaluator we make use of a slightly bigger unit of computation: a \emph{visit}.

@@

It is theoretically impossible to minimise the total number of visits performed for the full abstract syntax tree, because at compile-time we do not have a concrete abstract syntax tree at hand and only know about the grammar. We therefore try to minimise the maximum number of visits for any nonterminal, which is the number of alternating pairs of inherited and synthesized attributes in the total order.

We use the following algorithm:
\begin{enumerate}
\item Construct the initial SAT problem and solve
\item Repeat while there is a solution:
\begin{enumerate}
\item Find the longest chain of alternating inherited/synthesized attributes (which is the maximum number of visits)
\item If this maximum is lower then the previous best, store this result
\item Add a constraint stating that at least one of the edges in this chain must be reversed
\item Call solve again
\end{enumerate}
\end{enumerate}

Remark that removing the longest chain of visits might lead to a worst schedule, as this could result in other nonterminals using even more visits. We therefore need to store the best result found so far separately. However, it is guaranteed that this procedure will find the optimal result, and experiments have shown it usually does so quickly.

\subsection{Eager attributes} \label{sect:eagerattributes}
Another optimisation is the ability to define \emph{eager attributes}. Eager attributes are attributes that should be computed as soon as possible, and must be annotated by the attribute grammar programmer as such. We would like our scheduling algorithm then to schedule them as early as possible in the total order.

@@

We can use the following algorithm to implement this:
\begin{enumerate}
\item Construct the initial SAT problem and solve
\item Create a set $\mathcal{E}$ with the eager attributes
\item While $\mathcal{E}$ is not empty:
\begin{enumerate}
\item Take an attribute $\mathcal{A}$ from the set $\mathcal{E}$
\item For all dependencies of $\mathcal{A}$, add constraint that at least one the edges must be reverted
\item Solve, and if there is a solution:
\begin{enumerate}
\item Add $\mathcal{A}$ again to the set $\mathcal{E}$
\end{enumerate}
\item If there is no solution:
\begin{enumerate}
\item Revert to the previous solution (MiniSat allows this)
\item Add all dependencies of $\mathcal{A}$ to $\mathcal{E}$
\end{enumerate}
\end{enumerate}
\end{enumerate}

One important remark here is that $\mathcal{E}$ does not necessarily contain attributes from the same nonterminal. If at the root of the abstract syntax tree we would like to have an attribute depend on as few other attributes as possible, it may be the case that deeper in the tree the attributes need to be reordered in order to achieve that goal. The dependencies of $\mathcal{A}$ therefore means the dependencies in a production nonterminal graph, which can include attributes of other nonterminals (because children of a production can be of different nonterminals).

In contrast to the previous optimisation, this algorithm does not find a global optimum. Because optimistic choices are made this algorithm finds a local minimum, which may be much worse than the global optimum. We have not been able to verify what the influences of this are.
